Long term cloud enhancement events of global solar irradiance over Thessaloniki

### 2024-05-26 Article_B5/Article_B5_plain.tex
Commit 2024-05-27 00:14

  \title{Long term cloud enhancement events of global solar irradiance
over Thessaloniki}
    \author[LAP]{Athanasios N. Natsis%
  \corref{cor1}%
  \fnref{1}}
   \ead{natsisphysicist@gmail.com} 
    \author[LAP]{Alkiviadis Bais%
  %
  }
   \ead{abais@auth.gr} 
    \author[LAP]{Charikleia Meleti%
  %
  \fnref{2}}
   \ead{meleti@auth.gr} 
      \affiliation[LAP]{
    organization={Laboratory of Atmospheric Physics, Physics Department,
Aristotle University of
Thessaloniki},city={Thessaloniki},postcode={54124},country={Greece},}
    \affiliation[Another University]{
    organization={Department},addressline={A street
29},city={Manchester,},postcode={2054 NX},state={State},country={The
Netherlands},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{This is the first author footnote.}
    \fntext[2]{Another author footnote.}

  \begin{abstract}
  Here wii be the abstract.

  It consists of two paragraphs.
  \end{abstract}
    \begin{keyword}
    cloud enhancement \sep total solar radiation \sep global horizonta
irradiance \sep 
    over irradiance
  \end{keyword}

@@

Some of the proposed underling mechanisms of those events, have been
summarized by \citet{Gueymard2017}, and include scattering on the edge
of cumulus clouds or through thin cirrus. Further investigation with
radiation transfer modeling methods and observations have pointed as the
prevailing mechanism, the forward Mie scattering
\citep{Pecenak2016, Thuillier2013, Yordanov2013, Yordanov2015}, through
the clouds. The overall phenomenon depends on different interactive
factors, that include cloud thickness, constitution and type; and the
relative position of the sun, the clouds, and the observer
\citep{Gueymard2017, Veerman2022}. As such, there are multiple
contributing mechanism that are responsible for the observed irradiance
enhancements.

Cloud enhancements have been reported, to be able to exceed in intensity
even the value of the solar constant, resulting of clear indexes above
unit. A summary of extreme enchantment cases have been compiled by
\citet{Martins2022}. There are also some practical implication of the
cloud enchantments. The intensity and duration of enhancements can
effect the efficiency and stability of photovoltaic power production
\citep{Lappalainen2020, Jaervelae2020}, and extreme enhancements cases,
have the potential to compromise the integrity of photovoltaic plants
infrastructure \citep{DoNascimento2019}. It has also been demonstrated,
that these events can interfere in the comparison of ground data and
satellite observations \citep{Damiani2018}

@@

In this study, we evaluated the effects cloud enhancements on the total
downward radiation by studding the occurrences, their intensity, and
their duration in a thirty-year period at the city of Thessaloniki. We
used modeled clear sky irradiance, as a baseline to identify cloud
enhancements. We were able to determine some trends of the phenomenon,
it's Climatology, and some of their general characteristics. We weren't
able to find a comparable study that provides trends for similar long
term dataset, as ours. The recording of the radiation signal is the mean
of one-minute. Thus, the minimum resolution of cloud enhancement events
(CE), in this study is one minute.

In the relative bibliography different definition have been used for
these events, some of the are summarized by \citet{Gueymard2017}. Here,
we defined as cloud enhancement events (CE) the cases when the measured
global horizontal irradiance (GHI) at ground level, exceeds the expected
value under clear-sky conditions. Similar, we define as extreme cloud
enhancement events (ECE), the cases when GHI, exceeds the Total Solar
Irradiance on horizontal plane at ground level. Although the duration of
these bursts varies, from instantaneous to several minutes, here we are
constrained by the recorded data, to one-minute steps.

@@

To further improve the quality of the irradiance data, a manual
screening was performed to remove inconsistent and erroneous recordings
that can occur stochastically or systematically during the continuous
operation of the instruments. The manual screening was aided by a
radiation data quality assurance procedure, adjusted for the site, which
was based on the methods of Long and Shi~\citep{Long2006, Long2008a}.
Thus, problematic recordings have been excluded from further processing.
Although it is impossible to detect all false data, the large number of
available data, and the aggregation scheme we used, ensures the quality
of the radiation measurements used in this study. To preserve an
unbiased representation of the data we applied a constraint, similar the
one used by \citet{CastillejoCuberos2020}. Where, for each valid hour of
day, there must exist at least 45 minutes of valid measurements,
including nighttime near sunrise and sunset. Days with less than 5 valid
hours are rejected completely. Furthermore, due to the significant
measurement uncertainty when the Sun is near the horizon, and due to
some systematic obstructions by nearby buildings, we have excluded all
measurements with solar zenith angle (SZA) greater than \(78^\circ\).

@@

To be able to detect the CE cases, we had to establish a baseline, above
which we can characterize each data point as an enhancement event, by
estimating the occurring over irradiance (OIR). The OIR, here is defined
as the irradiance difference of the measured one-minute \(\text{GHI}_i\)
from the CE identification criterion in
Equation\nobreakspace\ref{eq:CE4}
(\(\text{OIR}_i = \text{GHI}_i - \text{GHI}_\text{CSlim}\)). To have an
estimation, and a first insight on the phenomenon, we experimented with
two simple approaches for the reference. The Haurwitz's model
\citep{Haurwitz1945}, which is as simple clear sky model, and we had
already adjusted and had good fit with our data \citep{Natsis2023}, and
the TSI at the top of the atmosphere. We have tested both cases by using
an appropriate relative threshold and/or an additional constant offset.
The initial results, showed that we can detect a big portion of the CE
events. These results were helpful, and helped to establish some
criteria to further improve the CE identification. It was evident, by
inspecting the daily plot of irradiance, that changes on the atmospheric
conditions introduced numerous false positive and false negative
results. To produce a more accurate reference, we had to take into
account more factors that effect the clear sky radiation. So we used a
radiation transfer model.

@@

We approximated the expected clear sky GHI by using a radiation transfer
model. The simulations was performed by the well established model
Libradtran \citep{Emde2016}, a similar approach, was also used by
\citet{Vamvakas2020} for creating a clear sky reference. Because of the
lack of observational data for the whole period, we used some long term
climatological data for the main factors responsible for the attenuation
of the broadband downward solar radiation in the atmosphere. Which are
mainly, the aerosols and the water vapors. Fortunately, our site
participates in the Aerosol Robotic Network (AERONET)
\citep{Giles2019, Buis1998}, as we operate a Cimel photometer since
2003, collocated with the CM-21 pyranometer. The mean monthly aerosol
optical depth (AOD) on different wavelengths is provided by AERONET,
along with the equivalent water column height in the atmosphere.

For completeness, we will describe here the main points of the radiation
simulation procedure. We used as input the spectrum of
\citet{Kurucz1994} in the range \(280\) to \(2500\,\text{nm}\), with the
Libradtran radiation transfer solver ``disort'' on a ``pseudospherical''
geometry and the ``LOWTRAN'' gas parameterization. For each combination
of conditions we use a SZA step of \(0.2^\circ\). For the atmospheric
characteristics, we iterated for combinations of AOD at
\(500\,\text{nm}\) (\(\tau_{500\text{nm}}\)) with additional offsets of
\(\pm1\) and \(\pm2\sigma\), and water column (\(wc\)), also with
offsets of \(\pm1\) and \(\pm2\sigma\). We applied them on two
atmospheric profiles, from the Air Force Geophysics Laboratory (AFGL).
The ``AFGL atmospheric constituent profile, midlatitude summer''
(afglms) and the ``AFGL atmospheric constituent profile, midlatitude
winter'' (afglmw) \citep{Anderson1986}.

To create a look-up table, that aligns with our dataset, we applied some
adjustments. To account for the Sun's variability, in our one-minute GHI
measurements, we adapted each modeled value by scaling the model's input
spectrum integral, to the corresponding total solar irradiance (TSI)
provided by NOAA \citep{Coddington2005}. Also, we account for the effect
of the Earth -- Sun distance on the irradiance, by using the distance
calculated by the Astropy \citep{AstropyCollaboration2022} software
library. As needed, we interpolate the resulting irradiances to the
exact SZA of our measurements. For each period of the year, we used the
appropriate atmospheric profile (afglms or afglmw). Finally, we
calculated the clear sky irradiance value at the horizontal plane. Thus,
we were able to emulate different atmospheric condition and levels of
atmospheric clearness for the climatological conditions of the site.
With this method, the modeled clear sky irradiances can be directly
compared to each measured one-minute value of GHI, for different
conditions of atmospheric clearness.

@@

The above clear sky reference values are not enough to describe the long
term variation, due to the changes of the atmospheric constituents. As
observed by \citet{Natsis2023}, there is a long term brightening effect,
over Thessaloniki for the studied period. In order to better describe
this change, we used, estimations of AOD change derived from two
sources. For 1993 -- 2005, \citet{Kazadzis2007} have found a change of
\(-3.8\,\%\) per year at AOD on \(340\,\text{nm}\), using a Brewer
spectrophotometer. For 2005 to 2023, we derived the trend from the
available AERONET data.

To create a unified trend for the long term change of the clear sky
irradiance, we simulated the values of AOD derived from those sources
with Libradtran. For both, we have used the AOD at \(500\,\text{nm}\),
which was inferred by the available Ångström coefficients, and we choose
the SZA of \(55^\circ\) as a representative value for all runs.

According to \citet{Kazadzis2007}, in the period 1997 -- 2005, the mean
AOD of \(340\,\text{nm}\) was calculated to \(0.403\) with the use of a
Brewer spectrophotometer, with a linear trend of \(-3.8±0.93\,\%\) per
year. This corresponds to a AOD \(340\,\text{nm}\) change of \(0.0153\)
per year. Using the alpha Ångström coefficient of \(\alpha = 1.6\), this
translates to a change in Ångström beta coefficient of \(\beta=0.00272\)
per year (or \(\beta=0.084\) in 1997 and \(\beta=0.059\) in 2005). By
interpolating the resulting clear sky irradiance, we have a trend of
\(+0.21\,\%\) per year for the clear sky irradiance. For the period 2005
-- 2005 we used the mean monthly values from AERONET, with a similar
simulation scheme. We computed the trend of clear sky radiation for each
month of the year, which gave us a mean clear sky radiation trend of
\(+0.14\,\%\) per year. We applied this long term change to the clear
sky irradiance, in order to create a more realistic representation of
the irradiance for the whole study period
(Figure\nobreakspace{}\ref{fig:CS-change}).

@@

\caption{Simulated long term change of clear sky irradiance relevant to its climatological values.}\label{fig:CS-change}
\end{figure}

@@

In the process of defining the above details, we implemented an
empirical method, by manual inspection of the CE identification, on
specially selected days from the whole dataset. We used seven sets of
selected days, with characteristics relevant to the efficiency of the
identification threshold. These sets were random groups of about 20 to
30 days with the following characteristics: (a) the largest over
irradiance CE events, (b) the largest daily total over irradiance, (c)
without clouds (by implementing a clear sky identification algorithm as
discussed in \citet{Natsis2023}), (d) without clouds and without EC
events, (e) with at least \(60\,\%\) of the day length without clouds
and some EC events, (h) random days and (i) some manual selected days
that were included during the manual inspection. Where it was needed,
for some of the edge cases, we also used images from a sky-cam, to
further aid the decisions of the manual inspection.

The definition of the CE events with this method, has a level of
subjectivity. The actual clear sky irradiance, is not known, and can
only be approximated. Although, because our main focus is to be able to
identify the OIR created by the clouds, we expect that we were able to
detect all the major events of OIR due to CE. Where, minor enhancements
may be not detected, are less significant for most of our results.

Another aspect of the CE events, that is often reported in the relative
bibliography, are cases of extreme cloud enhancement (ECE). These are
cases of CE, where the measured intensity of the irradiance, exceeds the
equivalent TSI on the top of the atmosphere, and satisfy the
Equation\nobreakspace{}\ref{eq:ECE}. \begin{equation}
\text{ECE}: \text{GHI}_\text{i} > \cos(\theta) \times E_{i\odot} / r_{i}^2
\label{eq:ECE}
\end{equation} where: \(\theta\) the solar zenith angle, \(E_{\odot}\)
the solar constant, \(r\) the Sun -- Earth distance, and \(i\) each of
the one-minute observation.

To give a better understanding of the CE identification procedure, we
provide an example of the CE identification for a selected day in the
Appendix (Figure\nobreakspace{}\ref{fig:example-day}), where the use of
the reference and the threshold are shown, along with the relevant
measurements. Similar we provide an example plot of one year data, where
the relation of the GHI and clear sky reference is shown, along with the
OIR intensity also in the Appendix
Figure\nobreakspace{}\ref{fig:example-year}.

@@

Our dataset, after the data selection processing, consists of 6144534
records of GHI, of which \(1.764\,\%\) are CE events and \(0.036\,\%\)
are ECE events. The highest GHI recorder was \(1416.6\,W/m^2\) on 24~May
2007. The absolute stronger CE event had an OIR of \(345.86\,W/m^2\) on
24~May 2007. The relative stronger CE event was \(49.7\,\%\) above the
clear sky threshold, on 28~October 2016.

\hypertarget{trends}{%
\subsection{Trends}\label{trends}}

@@

\hypertarget{ce-climatology}{%
\subsection{CE Climatology}\label{ce-climatology}}

Another interesting aspect of the CE cases occurrences, is their
seasonal cycle. In
Figure\nobreakspace{}\ref{fig:relative-month-occurancies}, we have a
monthly box plot (whisker plot), where the values have been normalized
by the highest median value, that occurs on June. Although the number of
occurrences has a wide spread throughout the study period, the most
active period of CE occurrences is during May and June. During the
Winter (December -- February) the CE cases are about \(25\,\%\) of the
maximum. For the rest months the occurrences gradually ramp between the
maximum and minimum. These seasonality is a combinations of the clouds
occurrences, and the relevant position of the sun during the year.
Unfortunately, the lack of detailed data of cloud formation, type and
location is not allowing further analysis.

@@

The distribution of the CE over irradiance intensity follows an
exponential decline Figure\nobreakspace{}\ref{fig:ovir-distribution},
where there is an inverse relation between the events frequency and
events intensity. This is expected as, the stronger the CE events
become, the rarer the particular atmospheric and sun conditions will
occur. This fact is indicative to the magnitude and probability of the
expected OIRs events over Thessaloniki. Similar distribution of CE
occurrences have been reported by \citet{Vamvakas2020}, where the
magnitude of OIR is higher due to the location of the the city of
Patras, \(2.5^\circ\) closer to the equator.

@@

\hypertarget{groups-stats}{%
\subsection{Groups stats}\label{groups-stats}}

In order to further study the characteristics of the CE events, we
grouped the single minute CE events to continuous CE groups (CEG). Thus,
a CEG consists of one or more successive CE cases, and can represent
cloud enhancement conditions spanning a duration of multiple minutes. We
have identified 28468 groups, where the group with the longest duration
lasted 140 minutes on 07~July 2013. By examining the frequency
distribution of the CEG durations
(Figure\nobreakspace{}\ref{fig:ceg-duration-distribution}), we can
conclude, that although we detected some CEG with durations longer than
an hour, about \(80\,\%\) of the CEG have a duration of less than 5
minutes.

@@

The relation between the duration and the mean OIR intensity of the
groups have been studied (Figure\nobreakspace{}\ref{fig:group-2d}). We
observe that the GCE events tend to have either long duration, or large
intensity, events with strong enhancement and long duration are very
rare. A similar result on this relation, have been also reported by
\citet{Zhang2018}, on a study using a far higher sampling rate than
ours.

@@

\caption{Relation of mean over irradiance and CE group duration}\label{fig:group-2d}
\end{figure}

\hypertarget{extreme-ce-above-tsi}{%
\subsection{Extreme CE above TSI}\label{extreme-ce-above-tsi}}

A characteristic of the CE events that is commonly reported and has some
significance on the solar energy production infrastructure are the
extreme CE events (ECE). Where solar irradiance exceeds the expected
irradiance on top of the atmosphere, we have defined this as ECE
(Equation\nobreakspace{}\ref{eq:ECE}).

Analogous to Figure\nobreakspace{}\ref{fig:relative-month-occurancies}
we have computed the distribution of the number of occurrences of ECE
events by month in
Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}. The most
active period for ECE events is in the spring and the start of the
summer (March -- June), followed by a period in the late fall (September
and October). This is probably an effect of the weather characteristics
of those periods where, there are continuous alternations between clear
sky periods and clouds.

@@

The distribution of the ECE events
(Figure\nobreakspace{}\ref{fig:P-extreme-distribution}), shows that
there are rare cases where the OIR can exceed the TSI even more than
\(400\,\text{W}/\text{m}^2\), with the \(75\,\%\) of the cases to be
below \(200\,\text{W}/\text{m}^2\). Similar to CE cases, the
distribution of ECE have a similar pattern with the results of
\citet{Vamvakas2020}, with the difference that the OIR values are lower
fro Thessaloniki.

@@

By creating a clear sky approximation of the GHI, which can represent
the long- and short-term variation of the expected clear sky GHI, we
were able to identify cases of CE events. By analyzing the CE cases we
found an increase of \(45.6\pm 21.9\,\text{cases}/\text{year}\), with
the mean annual total energy of the CE events increasing with a rate of
\(116.9\pm 67.8\,\text{kj}/\text{year}\). The most active season of CE
events over Thessaloniki is concentrated on early summer, on May and
June.

The magnitude of the ECE events doesn't exceeds the values reported from
sites with more favourable conditions for the phenomenon. Although, the
climatological characteristic of the ECE events, showed that the most
active months are spread on half of the year (March -- June and
September and October). We found that CE conditions, can have a duration
of more than an hour in rare cases, with the bulk of the cases having a
duration under 5 minutes. Some of the characteristics of CE and ECE
events we analysed, have strong similarities with results by
\citet{Vamvakas2020}, for a city southern of Thessaloniki, with
analogues differences, due to the higher intensity of the solar
radiation.

An interpretation of CE trends, shows that the interaction of GHI with
the clouds through this 30 year period, is a dynamic phenomenon that's
needs further investigation. Although, to approach it, we need more long
term observations of AOD and clouds characteristics.

>>>>

  \title{Long term cloud enhancement events of global solar irradiance
over Thessaloniki}
    \author[LAP]{Athanasios N. Natsis%
  \corref{cor1}%
  \fnref{1}}
   \ead{natsisphysicist@gmail.com} 
    \author[LAP]{Alkiviadis Bais%
  %
  }
   \ead{abais@auth.gr} 
    \author[LAP]{Charikleia Meleti%
  %
  \fnref{1}}
   \ead{meleti@auth.gr} 
      \affiliation[LAP]{
    organization={Laboratory of Atmospheric Physics, Physics Department,
Aristotle University of
Thessaloniki},city={Thessaloniki},postcode={54124},country={Greece},}
    \cortext[cor1]{Corresponding author}
    \fntext[1]{This is the first author footnote.}
    \fntext[2]{Another author footnote.}

  \begin{abstract}
  Here will be the abstract.
  \end{abstract}
    \begin{keyword}
    cloud enhancement \sep total solar radiation \sep global horizontal
irradiance \sep 
    over irradiance
  \end{keyword}

@@

Some of the proposed underling mechanisms of those enhancements, have
been summarized by \citet{Gueymard2017}, and include scattering on the
edge of cumulus clouds or through thin cirrus. Further investigation
with radiation transfer modeling methods and observations have pointed
as the prevailing mechanism, the forward Mie scattering
\citep{Pecenak2016, Thuillier2013, Yordanov2013, Yordanov2015}, through
the clouds. The overall phenomenon depends on different interactive
factors, which includes cloud thickness, constitution and type; and the
relative position of the sun and the clouds
\citep{Gueymard2017, Veerman2022}. As such, there are multiple
contributing mechanisms that are responsible for the observed irradiance
enhancements.

On multiple sites, cloud enhancements have been reported, to be able to
briefly exceed the value of the solar constant, resulting of clear
indexes above unit. A summary of extreme enchantment cases (ECE) have
been compiled by \citet{Martins2022}. There are also, some practical
implication of the cloud enchantments. The intensity and duration of
enhancements can effect the efficiency and stability of photovoltaic
power production \citep{Lappalainen2020, Jaervelae2020}, and extreme
enhancements cases, have the potential to compromise the integrity of
photovoltaic plants infrastructure \citep{DoNascimento2019}. It has also
been demonstrated, that these events can interfere in the comparison of
ground data and satellite observations \citep{Damiani2018}

@@

In this study, we evaluated the effects CE on the GHI by studding the
occurrences, their intensity, and their duration in a thirty-year period
at the city of Thessaloniki. We used modeled clear sky irradiance, as a
baseline to identify cloud enhancements. We were able to determine some
trends of the phenomenon, it's Climatology, and some of their general
characteristics. We weren't able to find a comparable study that
provides trends for similar long term dataset, as ours. The recording of
the GHI signal is the mean of one-minute. Thus, the minimum resolution
of CE events, in this study is one minute.

In the relative bibliography different definition have been used for
these events, some of the are summarized by \citet{Gueymard2017}. Here,
we defined as CE events, the cases when the measured GHI at ground
level, exceeds the expected value under clear-sky conditions. Similar,
we define as extreme cloud enhancement events (ECE), the cases when GHI,
exceeds the Total Solar Irradiance (TSI) on horizontal plane at ground
level. Although the duration of these bursts varies, from instantaneous
to several minutes, here we are constrained by the recorded data, to
one-minute steps.

@@

To further improve the quality of the irradiance data, a manual
screening was performed, in order to remove inconsistent and erroneous
recordings that can occur stochastically or systematically during the
long operation of the instrument. The manual screening was aided by a
radiation data quality assurance procedure, adjusted for the site, which
was based on the methods of Long and Shi~\citep{Long2006, Long2008a}.
Thus, problematic recordings have been excluded from further processing.
Although it is impossible to detect all false data, the large number of
available data, and the aggregation scheme we used, ensures the quality
of the radiation measurements used in this study. To preserve an
unbiased representation of the data we applied a constraint, similar the
one used by \citet{CastillejoCuberos2020}. Where, for each valid hour of
day, there must exist at least 45 minutes of valid measurements,
including nighttime, when near sunrise and sunset. Days with less than 5
valid hours were rejected completely. Furthermore, due to the
significant measurement uncertainty, when the Sun is near the horizon,
and due to some systematic obstructions by nearby buildings, we have
excluded all measurements with solar zenith angle (SZA) greater than
\(78^\circ\).

@@

To be able to detect the CE cases, we had to establish a baseline, above
which we characterized each data point as an enhancement event, by
estimating the occurring over irradiance (OIR). The OIR, is defined as
the irradiance difference of the measured one-minute \(\text{GHI}_i\)
from the CE identification criterion
(\(\text{OIR}_i = \text{GHI}_i - \text{GHI}_\text{CSlim}\)), as defined
in Equation\nobreakspace\ref{eq:CE4}. To have an estimation, and a first
insight on the phenomenon, we experimented with two simple approaches
for the reference. The Haurwitz's model \citep{Haurwitz1945}, which is
as simple clear sky model, and we had already adjusted and had good fit
with our data \citep{Natsis2023}, and the TSI at the top of the
atmosphere. We have tested both cases by using an appropriate relative
threshold and/or an additional constant offset. The initial results,
showed that we can detect a big portion of the CE events. These results
were helpful, as they are independent from unknown factors, and helped
us to establish some criteria to further improve the CE identification.
It was evident, by inspecting the daily plot of irradiance, that changes
on the atmospheric conditions introduced numerous false positive and
false negative results. To produce a more accurate reference, we had to
take into account more factors that effect the clear sky radiation. So
we used a radiation transfer model in order to include the effects of
ADO and the water vapors.

@@

We approximated the expected clear sky GHI by using the Libradtran
radiation transfer model \citep{Emde2016}, a similar approach, was also
used by \citet{Vamvakas2020} for creating a clear sky reference for
studding cloud enhancement events. Main factors responsible for the
attenuation of the broadband downward solar radiation in the atmosphere
are the aerosols and the water vapors. Because of the lack of
observational data for the whole period, we used some long term
climatological data to recreate the whole period. Fortunately, our site
participates in the Aerosol Robotic Network (AERONET)
\citep{Giles2019, Buis1998}, as there is in operation a Cimel photometer
since 2003, collocated with the CM-21 pyranometer. The mean monthly AOD
on different wavelengths are provided by AERONET, along with the
equivalent water column height in the atmosphere.

For completeness, we will describe here the main points of the radiation
simulation procedure. We used as input the spectrum of
\citet{Kurucz1994} in the range \(280\) to \(2500\,\text{nm}\), with the
Libradtran radiation transfer solver ``disort'' on a ``pseudospherical''
geometry and the ``LOWTRAN'' gas parameterization. For each combination
of conditions we use a SZA step of \(0.2^\circ\). For the atmospheric
characteristics, we iterated on combinations of AOD at
\(500\,\text{nm}\) (\(\tau_{500\text{nm}}\)) with additional offsets of
\(\pm1\) and \(\pm2\sigma\), and water column (\(w\)), also with offsets
of \(\pm1\) and \(\pm2\sigma\). We applied them on two atmospheric
profiles, from the Air Force Geophysics Laboratory (AFGL). The ``AFGL
atmospheric constituent profile, midlatitude summer'' (afglms) and the
``AFGL atmospheric constituent profile, midlatitude winter'' (afglmw)
\citep{Anderson1986}.

To create a look-up table, that aligns with our dataset, we applied some
adjustments. To account for the Sun's variability, in our one-minute GHI
measurements, we adapted each modeled value by scaling the model's input
spectrum integral, to the corresponding TSI, provided by NOAA
\citep{Coddington2005}. Also, we account for the effect of the Earth --
Sun distance on the irradiance, by using the distance calculated by the
Astropy \citep{AstropyCollaboration2022} software library. As needed, we
interpolate the resulting irradiances to the SZA of our measurements.
For each period of the year, we used the appropriate atmospheric profile
(afglms or afglmw). Finally, we calculated the clear sky irradiance
value at the horizontal plane. Thus, we were able to emulate different
atmospheric condition and levels of atmospheric clearness for the
climatological conditions of the site. With this method, the modeled
clear sky irradiances can be directly compared to each measured
one-minute value of GHI, for different conditions of atmospheric
clearness.

@@

The above clear sky reference values are not enough to describe the long
term variation, due to the changes of the atmospheric constituents,
mainly AOD. As observed by \citet{Natsis2023}, there is a long term
brightening effect, over Thessaloniki for the studied period. In order
to better incorporate this change, we used, estimations of AOD change
derived from two sources. For 1993 -- 2005, \citet{Kazadzis2007} have
found a change of \(-3.8\,\%\) per year at AOD on \(340\,\text{nm}\),
using a Brewer spectrophotometer. For 2005 to 2023, we derived the trend
from the available monthly climatological data of AERONET.

To create a unified trend for the long term change of the clear sky
irradiance, we simulated the values of AOD derived from those sources
with Libradtran. For both inputs, we used the AOD at \(500\,\text{nm}\),
which was inferred by the available Ångström coefficients. We choose the
SZA of \(55^\circ\) as a representative value for all the runs.

According to \citet{Kazadzis2007}, in the period 1997 -- 2005, the mean
AOD of \(340\,\text{nm}\) was calculated to \(0.403\) with the use of a
Brewer spectrophotometer, with a linear trend of \(-3.8±0.93\,\%\) per
year. This corresponds to a change of \(0.0153\) per year for the AOD of
\(340\,\text{nm}\) . Using the alpha Ångström coefficient of
\(\alpha = 1.6\), this translates to a change in Ångström beta
coefficient of \(\beta=0.00272\) per year (or \(\beta=0.084\) in 1997
and \(\beta=0.059\) in 2005). By interpolating the resulting clear sky
irradiance, we have a trend of \(+0.21\,\%\) per year for the clear sky
irradiance. For the period 2005 -- 2005 we used the mean monthly values
from AERONET, with a similar simulation scheme. We computed the trend of
clear sky radiation for each month of the year, which gave us a mean
clear sky radiation trend of \(+0.14\,\%\) per year. We applied this
long term change to the clear sky irradiance, in order to create a more
realistic representation of the irradiance for the whole study period
(Figure\nobreakspace{}\ref{fig:CS-change}).

@@

\caption{Simulated long term change of the clear sky irradiance relevant to its climatological values, due to changed of AOD.}\label{fig:CS-change}
\end{figure}

@@

In the process of defining the above details, we implemented an
empirical method, by manual inspection of the CE identification results,
on selected days from the whole dataset. We used seven sets of days,
with characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 to 30 days with the
following characteristics: (a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance, (c) absence of clouds (by
implementing a clear sky identification algorithm as discussed in
\citet{Natsis2023}), (d) absence of clouds and absence of EC events, (e)
with at least \(60\,\%\) of the day length without clouds and presence
of EC events, (h) random selected days and (i) some manual selected days
that were included during the manual inspection process. Where it was
needed, for some of the edge cases, we also used images from a sky-cam,
to further aid the decisions of the manual inspection.

The definition of the CE events with this method, has a level of
subjectivity. The actual clear sky irradiance, is not known, and can
only be approximated. Although, because our main focus is to be able to
identify the OIR created by the clouds, we expect that we were able to
detect all the major events of OIR due to CE. Where, some CE events,
with very low OIR, may be not detected, these are few and their
magnitude will not effect our results.

Another aspect of the CE events, that is often reported in the relative
bibliography, are cases of extreme cloud enhancement (ECE). These are
cases of CE, where the measured intensity of the irradiance, exceeds the
equivalent TSI on the top of the atmosphere, and detected according to
the Equation\nobreakspace{}\ref{eq:ECE}. \begin{equation}
\text{ECE}: \text{GHI}_\text{i} > \cos(\theta) \times E_{i\odot} / r_{i}^2
\label{eq:ECE}
\end{equation} where: \(\theta\) the solar zenith angle, \(E_{\odot}\)
the solar constant, \(r\) the Sun -- Earth distance, and \(i\) each of
the one-minute observation.

To give a better understanding of the CE identification procedure, we
provided an example of the CE identification for a selected day, in the
Appendix (Figure\nobreakspace{}\ref{fig:example-day}), where the use of
the clear sky reference and the identification threshold are shown,
along with the relevant measurements. Similar, we provided an example
plot of data for one year, where the relation of the GHI and clear sky
reference is shown, along with the OIR intensity (Appendix
Figure\nobreakspace{}\ref{fig:example-year}).

@@

Our dataset, after the data selection processing, consists of 6144534
records of GHI measurements, of which \(1.764\,\%\) were identified as
CE events and \(0.036\,\%\) as ECE events. The highest recorded GHI due
to CE was \(1416.6\,W/m^2\) on 24~May 2007, and had the maximum value of
OIR at \(345.9\,W/m^2\). The relative stronger CE event was \(49.7\,\%\)
above the clear sky threshold, on 28~October 2016.

\hypertarget{long-term-trends}{%
\subsection{Long term trends}\label{long-term-trends}}

@@

\hypertarget{cloud-enhancements-climatology}{%
\subsection{Cloud enhancements
climatology}\label{cloud-enhancements-climatology}}

Another interesting aspect of the CE cases occurrences, is their
seasonal cycle. In
Figure\nobreakspace{}\ref{fig:relative-month-occurancies}, we have the
monthly box plot (whisker plot), where the values have been normalized
by the highest median value, that occurs on June. Although the number of
occurrences has a wide spread throughout the study period, the most
active period of CE occurrences is during May and June. During the
Winter (December -- February) the CE cases are about \(25\,\%\) of the
maximum. For the rest months, the occurrences gradually ramp between the
maximum and minimum. This seasonality is a combinations of the clouds
occurrences, and the relevant position of the sun during each season of
the year. Unfortunately, the lack of detailed data on cloud formation,
type and location is not allowing further analysis.

@@

The distribution of the OIR intensity, follows an exponential decline
Figure\nobreakspace{}\ref{fig:ovir-distribution}, where there is an
inverse relation between the CE events frequency and their intensity.
This is expected, as the stronger the CE events become, the rarer the
particular atmospheric and sun conditions can occur. This fact, is
indicative to the magnitude and the probability of the expected OIRs
events over Thessaloniki. Similar distribution of CE occurrences have
been reported by \citet{Vamvakas2020}, where the magnitude of OIR were
higher due to the location of the city of Patras, \(2.5^\circ\) closer
to the equator.

@@

\hypertarget{groups-of-cloud-enhancement}{%
\subsection{Groups of cloud
enhancement}\label{groups-of-cloud-enhancement}}

In order to study the total duration of the CE events, we grouped the
one minute CE events to continuous CE groups (CEG). Thus, a CEG consists
of one or more successive CE cases, and can represent cloud enhancement
conditions spanning a duration of multiple minutes. We have identified
28468 groups of CE in the whole period of study, where the group with
the longest duration lasted 140 minutes on 07~July 2013. By examining
the frequency distribution of the CEG durations
(Figure\nobreakspace{}\ref{fig:ceg-duration-distribution}), we can
conclude, that although we detected some CEG with durations longer than
an hour, about \(80\,\%\) of the CEG have a duration of less than 5
minutes.

@@

The relation between the duration and the mean OIR intensity of the
groups have also been studied (Figure\nobreakspace{}\ref{fig:group-2d}).
We observed that the GCE events tend to have either long duration, or
large intensity. Events with strong enhancement and long duration are
very rare. Similar results on this relation, have been reported by
\citet{Zhang2018}, on a study using a far higher sampling rate than
ours.

@@

\caption{Relation of the mean over irradiance and duration of GCE, where the color scale donotes the frequency of the respected events.}\label{fig:group-2d}
\end{figure}

\hypertarget{extreme-cloud-enhancement-events}{%
\subsection{Extreme cloud enhancement
events}\label{extreme-cloud-enhancement-events}}

An aspect of the CE events that is commonly reported and has some
significance on the solar energy production infrastructure are the
extreme CE events (ECE). Where solar irradiance exceeds the expected
irradiance on top of the atmosphere
(Equation\nobreakspace{}\ref{eq:ECE}). Analogous to
Figure\nobreakspace{}\ref{fig:relative-month-occurancies} we have
computed the distribution of the number of occurrences of ECE events by
month in Figure\nobreakspace{}\ref{fig:relative-month-occurancies-ECE}.
The most active period for ECE events is in the spring and the start of
the summer (March -- June), followed by a period in the late fall
(September and October). This is probably related to the weather
characteristics of these periods, where there are continuous
alternations between clear sky periods and clouds.

@@

The distribution of the ECE events
(Figure\nobreakspace{}\ref{fig:P-extreme-distribution}), shows that
there are rare cases where the OIR can exceed the TSI even more than
\(400\,\text{W}/\text{m}^2\), with the \(75\,\%\) of the cases to be
below \(200\,\text{W}/\text{m}^2\). Those finds are in accordance with
results form \citet{Vamvakas2020}, with the difference that the OIR
values are lower for Thessaloniki.

@@

By creating a clear sky approximation of the GHI, which represents the
long- and short-term variation of the expected clear sky GHI, we were
able to identify cases of CE events. After analyzing the CE cases, we
found an increase of \(45.6\pm 21.9\,\text{cases}/\text{year}\), with
the mean annual total energy of the CE events increasing with a rate of
\(116.9\pm 67.8\,\text{kj}/\text{year}\). The most active season of CE
events over Thessaloniki is concentrated on early summer, on May and
June.

The magnitude of the ECE events doesn't exceeds the values reported from
sites with more favourable conditions for the phenomenon. Although, the
climatological characteristic of the ECE events, showed that the most
active months are spread on half of the year (March -- June and
September and October). We found that CE conditions, can have a duration
of more than an hour in rare cases, with the bulk of the cases having a
duration under 5 minutes. Some of the characteristics of CE and ECE
events we analysed, have strong similarities with results by
\citet{Vamvakas2020}, for a city southern of Thessaloniki, with the
analogues differences on the intensity of the solar radiation.

An interpretation of the CE trends, shows that the interaction of GHI
with the clouds, through this 30 year period, is a dynamic phenomenon
that's needs further investigation. Although, to approach it, we need
more long term observations of AOD and clouds characteristics.
